---
title: "Problem Set 3 - Statistics"
format: html
editor: visual
---

## Problem Set 3

setting things up

```{r}
rm(list = ls())
library(here)
library(tidyverse)
library(ggfortify)
```

loading in the data

```{r}
bats <- read.csv(here("data", "bat_data.csv"))
```

clean up data

```{r}
bats$Species <- as.factor(bats$Species)
bats$age <- as.factor(bats$age)
bats$sex <- as.factor (bats$sex)
```

Does body mass differ between these five species of bats and, if so, how does body\
mass differ between the species?

```{r}
#following the steps for an analysis we must start with a plot
ggplot(bats, aes(x= Species, y= mass_g, fill = Species)) +
  geom_bar(stat = "identity")
```

From this, it can be hypothesized that there is going to be a statistical significance between body mass and species. The LACI species seems to have the highest mean body mass around over 600g while the PESU species has a mean mass of 150g.

```{r}
#check the assumptions.
ggplot(bats, aes(x= Species, y= mass_g)) +
  geom_boxplot() + geom_point() + theme_bw()
```

```{r}
#contruct the anova , the predictor variable is categorical and the response is a ratio data
model.bat <- lm(mass_g ~ Species, data = bats )
autoplot(model.bat)
```

```{r}
#looking at the output 
anova(model.bat)
```

From the output of the anova test, there is a significant difference difference between species and the mass. The p value is significantly less than .05. Body mass is highest in the LACI species and lowest in the PESU species. ( df= 4, F= 792.9, p= \< .05) .

Does body length differ between these five species of bats and, if so, how does body\
length differ between the species?

```{r}
#plot the data 
ggplot(bats, aes(x= Species, y= body_length_mm, fill = Species)) +
  geom_bar(stat = "identity")
```

```{r}
#another plot for the data 
ggplot(bats, aes(x= Species, y= body_length_mm)) +
  geom_boxplot() + geom_point() + theme_bw()
```

```{r}
#constructing the anova 
model.length <- lm(body_length_mm ~ Species, data = bats )
autoplot(model.length)
```

```{r}
#output of the anova
anova(model.length)
```

From the output of the anova test, there is a significant difference between species and the body length. The p value is less than 0 and significantly less than .05 so there is a difference between species. There is also a large f value which means high variance between groups than compared to within the group. ( df= 4, F= 70.498, p= \< .05)

Is the number of ticks found on the bats associated with their sex or age?

```{r}
#plot the data
ggplot(bats, aes(x= sex, y= tick_num)) +
  geom_boxplot() + geom_point() + theme_bw()
```

```{r}
#construct the t test because their is two groups to compare 
model.tick <- lm(tick_num ~ sex, data = bats )
autoplot(model.tick)
```

```{r}
#run the t test 
t.test(tick_num ~ sex, data = bats)
```

There is a significant difference between male and females and ticks which means ticks are associated with the sex of bats (t= -10.61, df= 117.45, p= \< .05 ) .

```{r}
#plot for age 
ggplot(bats, aes(x= age, y= tick_num)) +
  geom_boxplot() + geom_point() + theme_bw()
```

```{r}
#construct model for age 
model.tick <- lm(tick_num ~ age, data = bats )
autoplot(model.tick)
```

```{r}
#output of test for age 
t.test(tick_num ~ age, data = bats)
```

There was not a significant difference between the ages of bats and the number of ticks with the p vale of 0.1 which is greater than .05 ( t= 1.629, df=122, p= 0.1059) . This means amount of ticks do not associate with the age of bats.

Disregarding species, is there a relationship in bats between tail length and body length?

```{r}
ggplot(bats, aes(x= tail_length_mm , y= body_length_mm )) + geom_point() + theme_bw()
```

```{r}
model_btl <- lm(body_length_mm ~ tail_length_mm , data = bats)
autoplot(model_btl, smooth.colour = NA)
```

From this, there seems to be a positive linear relationship between tail length and body length.
